---
title: "eds213 Data Cleaning"
format: html
author: "Megan Hessel"
warning: false
---

Load Data

```{r}
#| output: false

# Libraries
library(tidyverse)

# file names
datadir_raw <- file.path("data","raw")

datadir_processed <- file.path("data", "processed/")

species_file <- "ASDN_Daily_species.csv"

snowsurvey_file <- "ASDN_Snow_survey.csv"

snowsurvey_csv <- read_csv(file.path(datadir_raw, snowsurvey_file))
snowsurvey_csv
```

## Clean `Water_cover` Column

### Dealing with non-numeric values

Look at the non-numeric values

```{r}
# Investigate the non numeric values 
snowsurvey_csv %>% 
  count(Water_cover) %>% 
  filter(is.na(as.numeric(Water_cover)))
```

```{r}
# Change all the non-numeric values 
snowsurvey_fixed <- snowsurvey_csv %>% 
  mutate(Water_cover = ifelse(Water_cover == "-", NA, Water_cover), 
         Water_cover = ifelse(Water_cover == ".", NA, Water_cover),
         Water_cover = ifelse(Water_cover == "n/a", NA, Water_cover),
         Water_cover = ifelse(Water_cover == "unk", NA, Water_cover)
         )

# Check if it worked
snowsurvey_fixed %>% 
  count(Water_cover) %>% 
  filter(is.na(as.numeric(Water_cover)))
```

```{r}
# Make the column numeric 
snowsurvey_fixed$Water_cover <- as.numeric(snowsurvey_fixed$Water_cover)
```

### Value Ranges

The water cover must range from 0 to 100. Let's look if there are any values outside that range.

```{r}
unique(snowsurvey_fixed$Water_cover)

# Check values greater than 0 
snowsurvey_fixed %>% 
  filter(Water_cover > 100)

# Check for negative values 
snowsurvey_fixed %>% 
  filter(Water_cover < 0)
```

There is a water cover value that is greater than 100 (353%) which is impossible! Need to fix it!

```{r}
# Change the values greater than 100 to NAs 
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Water_cover = ifelse(Water_cover > 100, NA, Water_cover))
```

## Clean the `Land_cover` column

### Dealing with non-numeric values

Look at the non-numeric values

```{r}
# Investigate non numeric values 
snowsurvey_fixed %>% 
  count(Land_cover) %>% 
  filter(is.na(as.numeric(Land_cover)))
```

Change non-numeric values to NA

```{r}
# Change non numeric values 
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Land_cover = ifelse(Land_cover == "-", NA, Land_cover), 
         Land_cover = ifelse(Land_cover == ".", NA, Land_cover), 
         Land_cover = ifelse(Land_cover == "n/a", NA, Land_cover), 
         Land_cover = ifelse(Land_cover == "unk", NA, Land_cover))

# Check if it worked 
snowsurvey_fixed %>% 
  count(Land_cover) %>% 
  filter(is.na(as.numeric(Land_cover)))

```

Convert column to numeric

```{r}
# Change Land_cover to numeric 
snowsurvey_fixed$Land_cover <- as.numeric(snowsurvey_fixed$Land_cover)
```

### Value Ranges

Investigate which value are greater than 100 or negative.

```{r}
# Values greater than 100 
snowsurvey_fixed %>% 
  filter(Land_cover > 100)

# Negative values 
snowsurvey_fixed %>% 
  filter(Land_cover < 0)
```

Convert impossible values to NAs

```{r}
# Convert negative values to NAs 
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Land_cover = ifelse(Land_cover < 0, NA, Land_cover))
```

## Clean Snow Cover column

We did this in class: non-numeric and impossible values

```{r}
# Fix the non-numeric values 
snowsurvey_fixed <- snowsurvey_fixed %>%
  mutate(Snow_cover = ifelse(Snow_cover == ".", NA, Snow_cover), 
         Snow_cover = ifelse(Snow_cover == "-", NA, Snow_cover), 
         Snow_cover = ifelse(Snow_cover == "n/a", NA, Snow_cover), 
         Snow_cover = ifelse(Snow_cover == "unk", NA, Snow_cover), 
         Snow_cover = ifelse(Snow_cover == "<1", 0, Snow_cover))

# Make the column numeric 
snowsurvey_fixed$Snow_cover <- as.numeric(snowsurvey_fixed$Snow_cover)

# Chnage negative, impossible values to NA 
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Snow_cover = ifelse(Snow_cover > 100, NA, Snow_cover))

```

## Total Cover

Recompute the `Total_cover` to ensure 3 snow, water, and land cover value

```{r}

# Recalculate the Total Cover = Sum of snow, water, and land cover
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Total_cover = rowSums(pick(Snow_cover, Water_cover, Land_cover), na.rm = TRUE))

# Look at all total cover values and make sure they are 0 (NAs in rows) or 100
snowsurvey_fixed$Total_cover

```
## Export cleaned snowsurvey as CSV

```{r}
# Check if the folder exists
dir.create(datadir_processed, showWarnings = FALSE)

# Save file as CSV 
write_csv(snowsurvey_fixed, file.path(datadir_processed, "all_cover_fixed_MEGAN_HESSEL.csv"))
```
